---
title: "The Leslie Matrix - revisited"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are told that annually:

- Each breeding adult will produce, on average, 600 eggs
- 5% of eggs will survive to become fry
- 20% of fry will survive to become juveniles
- 16% of juveniles will survive to become adults
- Adult survival rate year to year is 90%

Recall the Leslie matrix from EDS 212. Matrix multiplication (using the dot product `%*%` of the matrix with the populations at each life stage) yields the population structure at the following year. 

### Create the Leslie matrix:

```{r}
# Make the Leslie Matrix:
fish_leslie <- matrix(c(0, 0, 0, 600, 0.05, 0, 0, 0, 0, 0.20, 0, 0, 0, 0, 0.16, 0.9), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out: 
fish_leslie
```

The initial population structure is given by: 
```{r}
# Initial population structure (Roe, Fry, Juvenile, Adult): 
fish_year0 <- c(5000, 8000, 600, 400)
```


Add line-by-line comments wheres prompted in the code below, explaining what each is doing. Make sure that you are talking with your neighbors about what is really being created at each step. 

```{r}
# Model projections

# ------ WHAT HAPPENS HERE? ------ #
# ------ comment line-by-line ---- #



# The line below is setting time_yr as a sequence starting at 0 ending at 8, going up in increments of 1. 
time_yr <- seq(from = 0, to = 8, by = 1)

time_yr

# The line below is storing a matrix the size of the fish_year0 and size of time_yr which is a 9 by 4 matrix. 
proj_year <- matrix(ncol = length(fish_year0), nrow = length(time_yr))

proj_year

# The line below is storing everything in the first row of the new matrix as the initial population structure any term left empty will be fully populated.  
proj_year[1, ] <- fish_year0



# The for loop below starts at 2, and iterates for every term in the ith column of the proj_year, multiplies fish_leslie using the dot product by the year prior (proj_year[i-1]).
# not starting to populate at row one, so starting at 2 up to the length of time_yr, will take the dot product between system of linear equations
# i want to iterate where to find the population projection, dot product between matrix of coefficients and the row containing the vector of the population of the previous cycle. starting to populate at row 2. over these 8 years, for each of these populatoin stages, basd on the model this tells us what the population size is expected.


for (i in 2:length(time_yr)) {
  proj_year[i,] <- fish_leslie %*% proj_year[i-1,]
}


# The rest is wrangling & visualization (run to check it out): reassigning column names
colnames(proj_year) <- c("eggs", "fry", "juvenile", "adult")

#combining time year vector with the matrix of population projections
#melting this in long format
# one variable lifestage split over four columns to one column.pivot longer across all columns except time_yr, take all column nanmes and change it to lifestage it and take values and align them to stage_population

proj_df <- data.frame(time_yr, proj_year) %>% 
  pivot_longer(cols = -time_yr, names_to = "lifestage", values_to = "stage_population")

ggplot(data = proj_df, aes(x = time_yr, y = stage_population)) +
  geom_line(aes(color = lifestage)) +
  scale_y_log10()

```

Test different combinations of survival / reproduction and starting population structures to see what happens! 
```{r}
demo_longer <- demo %>% 
  pivot_longer(cols=c(cats, dogs), names_to= "species", values_to= "count")

demo <- data.frame(year=c(1995,2000,2005),
                   cats = c(2,4,6),
                   dogs =c(10, 14, 12))
demo_longer

```

